Linting Linearity in Core/System FC

### 2023-01-24 report.tex
Small improvements

Linear type systems~\cite{} add expressiveness to existing type systems by
stating that linear resources must be used \emph{exactly once}. In programming
languages with a linear type system, not using certain resources or using them
twice are type errors. Linear types can thus be used to, for example, guarantee
that file handles, socket descriptors, and allocated memory are freed exactly
once (leaks and double-frees become type errors)~\cite{}, implement safe
high-performance language interoperability~\cite{}, guarantee that quantum
entangled states are not duplicated~\cite{}, handle mutable state safely~\cite{},
something on session types guarantees~\cite{}, and much more.

% TODO: Also include bit about how it makes the basis of other type systems such
% as rusts owneship. How ?

@@

Under the lens of a linear type system, consider the variable $p$ to be a linear
resource created by the \texttt{malloc} keyword. Because $p$ is a linear
variable, it must be used exactly once. However, we are using it in two
different calls to \texttt{free}. With our new linearity assumptions, the
% TODO:
example program \emph{would not typecheck}! Some conclusion on linearity
goodness. In section~\ref{linear-types} we describe the fundamentals of linear
types in more detail with a formal definition and many examples.

@@

Linearity in Haskell stands out % from linearity in Rust and Idris2
for the two following distinct reasons:

@@

    \item Linear types permeate Haskell down to (its) \textbf{Core}, the
        intermediate language Haskell is desugared to. \textbf{Core} is a very
        small explicitly typed functional language to which multiple
        Core-to-Core optimizing transformations are applied in succession. The
        final Core program can be typechecked very fast (because of the explicit
        type annotations) and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimized
        Core program doesn't typecheck, something \emph{very wrong} has happened
        in a transformation!  We discuss Core in detail in
        section~\ref{core-section}.
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?
\end{itemize}

@@

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions that receive as an argument a handle to work
with and are responsible for closing it, we try to understand how the examples
could be disregarded as incorrect, and work our way up to linear types from
them. The first function ignores the received file handle and returns $\star$
(read unit), which is equivalent to C's \texttt{void}.
% TODO: Better code formatting
\begin{mathpar}
    \lambda h.~\textrm{return}~\star;
    \and
    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
\end{mathpar}
% TODO: Equation references?

Ignoring the file handle which should have been closed by the function makes the
first function incorrect. Similarly, the second function receives the file handle and
closes it twice, which is incorrect not because it falls short of the
specification, but rather because the program will crash while executing it.
%
Additionally, both functions share the same type, $\texttt{Handle} \to \star$, i.e.
a function that takes a \texttt{Handle} and returns $\star$. The second function also
shares this type because \texttt{close} has type $\texttt{Handle} \to \star$.
%
Under a simple type system such as C's both functions are type correct (the
compiler will happily succeed), but both have erroneous behaviours. The first
forgets to close the handle and the second closes it twice. Our goal is now to
reach a type system that rejects these two programs.

@@

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus\todo{citar o girard
  provavelmente ou o Barber (\url{https://www.lfcs.inf.ed.ac.uk/reports/96/ECS-LFCS-96-347/})}, a very simple language with linear
types, by defining what are syntatically valid programs through the grammar
in Fig.~\ref{fig:llcgrammar} and what programs are well typed through the typing
rules in Fig.~\ref{fig:llcrules}. The language features functions and function application ($\lolli$), two
flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
disjunction operator ($\oplus$)\todo{sum types} and the $\bang$
modality operator\todo{que serve para}. A typing
judgement for the linearly typed lambda calculus has the form
%
\[ \Gamma; \Delta \vdash M : A \]
%
where $\Gamma$ is the context of resources that may be used unrestrictedly, that
is, any number of times, $\Delta$ is the context of resources that must be used
linearly (\emph{exactly once}), $M$ is the program to type and $A$ is its type.
When resources from the linear context are used, they are removed from the
context and no longer available, and all resources in the linear context must be
used exactly once.

@@

Finally, the "of-course" operator ($\bang$) can be used to construct a resource that can be
used unrestrictedly ($\bang M$). Its introduction rule ($\bang I$) states that
to construct this resource means to add a resource to the unrestricted context,
which can then be used freely. To construct an unrestricted value, however, the
linear context \emph{must be empty} -- an unrestricted value can only be constructed if
it does not depend on any linear resource.
%
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\]
%
To utilize an unrestricted value $M$, we must
bind it to $u$ with $\llet{\bang u = M}{N}$ and can then use it in $N$
unrestrictedly, because $u$ extends the unrestricted context rather than the
linear context as we have seen thus far.

In section~\ref{sec:linear-haskell} we describe how linear types are defined in
Haskell, a programming language more featureful than the linearly typed lambda
calculus. We will see that, the theoretical principles underlying the linear
lambda calculus and linear Haskell are the same, and by studying them in this
minimal setting we can understand them at large.

@@

Haskell is a large feature-rich language but its relatively small core is based on a typed lambda
calculus. As such, there exist no statements and computation is done simply
through the evaluation of functions. Besides functions, one can define types and
their constructors and pattern match on said constructors. Function application
is denoted by an empty space\todo{juxtaposition} (\texttt{f~a} means \texttt{f} applied to
\texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
by the enumerated alternatives. All variable names start with lower case and
types start with upper case (excluding type variables). To make explicit the
type of an expression, the \texttt{::} operator is used (e.g.
$\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has type} function
from \texttt{Int} to \texttt{Bool}).

@@

    \item Local equality constraints and existential types by using GADTs, we
        explain ahead in more detail. A design for first class existential types
        with bi-directional type inference in Haskell has been published
        in~\cite{10.1145/3473569}, despite not being yet implemented in GHC.

@@

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are an advanced Haskell feature
which allows users to define data types as they would common algebraic
data types, with the added abilitiy to give explicit type signatures for the
data constructors where the result type may differ in the type parameters across
branches.
\todo{nao ha branches, é nos varios construtores. This allows for
  additional type invariants to be represented with GADTs through
  their type parameters, which restricts the use of specific
  constructors and their subsequent deconstruction through pattern
  matching}
% TODO: Rewrite this bit below
Pattern matching against GADTs can introduce local type refinements, that is,
pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop the
length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

To define the safe \texttt{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, making a call to \texttt{head} on a list of type
\texttt{Vec Z a} (an empty list) is a compile-time type error.

@@

The introduction of linear types to Haskell's type system is originally
described in Linear Haskell\cite{cite:linearhaskell}. While
in~\ref{sec:related-work-linear-haskell} we discuss the reasoning and design
choices behind retrofitting linear types to Haskell, here we focus on
linear types solely as they exist in the language, and re-work the file
handle example seen in the previous section along the way.
% TODO: useless information but the term "Linear Haskell" is sometimes also used
% to refer to the existence of linear types in Haskell in general.

@@

In Haskell, linear types are introduced through \emph{linearity on the function
arrow}, rather than through \emph{linearity in the kinds}. In practice, this
means function types are annotated with a linearity that defines whether a
function argument must be consumed \emph{exactly once} or whether it can be
consumed \emph{unrestrictedly} (many times). Additionally, algebraic data types
constructor consume its arguments linearly by default, but the constructor can
be specified to receive unrestricted arguments which, in contrast, do not need
to be consumed exactly once when pattern matched on.
%
As an example, consider from the following functions the first, which doesn't
typecheck because it is a linear function (annotated with the linearity of
\texttt{1}) that consumes its argument more than once, and the second, which is
an unrestricted function (annoted with \texttt{Many}) that typechecks because
its type allows the argument to be consumed unrestrictedly.

\begin{code}
    f :: a %1 -> (a, a)
    f x = (x, x)
\end{code}
\begin{code}
    g :: a %Many -> (a, a)
    g x = (x, x)
\end{code}

The function annotated with the linearity annotation of \texttt{1} is equivalent
to the linear function type ($\lolli$) seen in the linear lambda
calculs~\ref{sec:linear-types}. To encode the multiplicative linear pair
($\tensor$) we must create a pair data type with two linear components. To
consume an algebraic data type is to consume all its linear components once, so
to consume said pair data type we need to consume both its linear components,
which successfully encodes the multiplicative pair elimination rule ($\tensor
E$). To construct said pair data type we must provide two linear elements, each
consuming some required resources to be constructed, which encodes the
multiplicative pair introduction rule ($\tensor I$).

@@

The linearity annotations are a specialization of the more general so-called
\emph{multiplicity annotations}. A multiplicity of \texttt{1} entails that the
function argument must be consumed once, and a function annotated with it
($\to_1$) is called a linear function (also known as $\lolli$). A function with
multiplicity of \texttt{Many} ($\to_\omega$) allows the function argument to
be consumed unrestrictedly. Unrestricted functions are equivalent to the common
function type and, in fact, the usual function arrow ($\to$) implicitly has
multiplicity \texttt{Many}.  Multiplicities also smoothly allow for
\emph{multiplicity polymorphism}, which we describe below.

\begin{itemize}
    \item Linear Haskell definition
    \item Consuming values precisely
    \item Multiplicities
    \item Multiplicity polymorphism
    \item Que relaciona o systemFC com o calculo lambda linear
\end{itemize}

>>>>

Linear type systems~\cite{} add expressiveness to existing type systems by
stating that linear resources must be used \emph{exactly once}. In programming
languages with a linear type system, not using certain resources or using them
twice are type errors. Linear types can thus be used to, for example, guarantee
that file handles, socket descriptors, and allocated memory are freed exactly
once (leaks and double-frees become type errors)~\cite{}, implement safe
high-performance language interoperability~\cite{}, guarantee that quantum
entangled states are not duplicated~\cite{}, something on session types
guarantees~\cite{}, and much more.
% handle mutable state safely~\cite{}

@@

Under the lens of a linear type system, consider the variable $p$ to be a linear
resource created by the \texttt{malloc} keyword. Because $p$ is a linear
variable, it must be used exactly once. However, we are using it in two
different calls to \texttt{free}. With our new linearity assumptions, the
% TODO:
example program \emph{would not typecheck}! \todo{Some conclusion on linearity
goodness.} In section~\ref{linear-types} we describe the fundamentals of linear
types in more detail with a formal definition and many examples.

@@

Linearity in Haskell stands out % from linearity in Rust and Idris2
two distinct reasons:

@@

    \item Linear types permeate Haskell down to (its) \textbf{Core}, the
        intermediate language Haskell is desugared to. \textbf{Core} is a very
        small explicitly typed functional language to which multiple
        Core-to-Core optimizing transformations are applied in succession. The
        final Core program can be typechecked very fast (because of the explicit
        type annotations) and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimized
        Core program doesn't typecheck, something \emph{very wrong} has happened
        in a transformation!  We discuss Core in detail in
        section~\ref{core-section}.\todo{Mention systemFC}
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?
\end{itemize}

@@

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions that receive a handle to work with (that must be
closed before returning), we explore how the examples could be disregarded as
incorrect, and work our way up to linear types from them. The first function
ignores the received file handle and returns $\star$ (read unit), which is
equivalent to C's \texttt{void}.
%
\begin{mathpar}
    \lambda h.~\mathsf{return}~\star;
    \and
    \lambda h.~\mathsf{close}~h;~\mathsf{close}~h;
\end{mathpar}
% TODO: Equation references?

Ignoring the file handle which should have been closed by the function makes the
first function incorrect. Similarly, the second function receives the file
handle and closes it twice, which is incorrect not because it falls short of the
specification, but rather because the program will crash while executing it.
%
Additionally, both functions share the same type, $\texttt{Handle} \to \star$,
i.e.  a function that takes a \texttt{Handle} and returns $\star$. The second
function also shares this type because \texttt{close} has type $\texttt{Handle}
\to \star$.
%
Under a simple type system such as C's, both functions are type correct (the
compiler will happily succeed), but both have erroneous behaviours. The first
forgets to close the handle and the second closes it twice. Our goal is to reach
a type system that rejects these two programs.

@@

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus~\cite{}\todo{citar o girard
provavelmente ou o Barber
(\url{https://www.lfcs.inf.ed.ac.uk/reports/96/ECS-LFCS-96-347/})}, a very
simple language with linear types, by defining what are syntatically valid
programs through the grammar in Fig.~\ref{fig:llcgrammar} and what programs are
well typed through the typing rules in Fig.~\ref{fig:llcrules}. The language
features functions and function application ($\lolli$), two flavours of pairs,
additive ($\with$) and multiplicative ($\tensor$), a disjunction operator
($\oplus$) to construct sum types, and the $\bang$ modality operator which
constructs an unrestricted type from a linear one, allowing values inhabiting
$\bang A$ to be consumed unrestrictedly. A typing judgement for the linearly typed
lambda calculus has the form
%
\[ \Gamma; \Delta \vdash M : A \]
%
where $\Gamma$ is the context of resources that may be used unrestrictedly, that
is, any number of times, $\Delta$ is the context of resources that must be used
linearly (\emph{exactly once}), $M$ is the program to type and $A$ is its type.
When resources from the linear context are used, they are removed from the
context and no longer available, and all resources in the linear context must be
used exactly once.

@@

Finally, the "of-course" operator ($\bang$) can be used to construct a resource that can be
used unrestrictedly ($\bang M$). Its introduction rule ($\bang I$) states that
to construct this resource means to add a resource to the unrestricted context,
which can then be used freely. To construct an unrestricted value, however, the
linear context \emph{must be empty} -- an unrestricted value can only be constructed if
it does not depend on any linear resource.
%
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\]
%
To utilize an unrestricted value $M$, we must bind it to $u$ with $\llet{\bang u
= M}{N}$ which can then be used in $N$ unrestrictedly, because $u$ extends the
unrestricted context rather than the linear context as we have seen thus far.

In section~\ref{sec:linear-haskell} we describe how linear types are defined in
Haskell, a programming language more featureful than the linearly typed lambda
calculus. We will see that the theoretical principles underlying the linear
lambda calculus and linear Haskell are the same, and by studying them in this
minimal setting we can understand them at large.

@@

Haskell is a large feature-rich language but its relatively small core is based
on a typed lambda calculus. As such, there exist no statements and computation
is done simply through the evaluation of functions. Besides functions, one can
define types and their constructors and pattern match on said constructors.
Function application is denoted by the juxtaposition of the function expression
and its arguments, which often means empty space between terms (\texttt{f~a}
means \texttt{f} applied to \texttt{a}). Pattern matching is done with the
\texttt{case} keyword followed by the enumerated alternatives. All variable
names start with lower case and types start with upper case (excluding type
variables). To make explicit the type of an expression, the \texttt{::} operator
is used (e.g.  $\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has
type} function from \texttt{Int} to \texttt{Bool}).

@@

    \item Local equality constraints and existential types by using GADTs, which
        we explain ahead in more detail. A design for first class existential
        types with bi-directional type inference in Haskell has been published
        in~\cite{10.1145/3473569}, despite not being yet implemented in GHC.

@@

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are an advanced Haskell feature
that allows users to define data types as they would common algebraic data
types, with the added abilitiy to give explicit type signatures for the data
constructors where the result type may differ in the type parameters (e.g., we
might have two constructors of the same data type \texttt{T~a} return values of
type \texttt{T~Bool} and \texttt{T~Int}).
%
This allows for additional type invariants to be represented with GADTs through
their type parameters, which restricts the use of specific constructors and
their subsequent deconstruction through pattern matching.
%
% TODO: Rewrite this bit below
Pattern matching against GADTs can introduce local type refinements, that is,
pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop the
length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

To define the safe \texttt{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, making a call to \texttt{head} on a list of type
\texttt{Vec Z a} (an empty list) a compile-time type error.

@@

The introduction of linear types to Haskell's type system is originally
described in Linear Haskell\cite{cite:linearhaskell}. While
in~\ref{sec:related-work-linear-haskell} we discuss the reasoning and design
choices behind retrofitting linear types to Haskell, here we focus on
linear types solely as they exist in the language, and re-work the file
handle example seen in the previous section to make sure it doesn't typecheck.
% TODO: useless information but the term "Linear Haskell" is sometimes also used
% to refer to the existence of linear types in Haskell in general.

@@

In Haskell, linear types are introduced through \emph{linearity on the function
arrow}, rather than through \emph{linearity in the kinds}. In practice, this
means function types are annotated with a linearity that defines whether a
function argument must be consumed \emph{exactly once} or whether it can be
consumed \emph{unrestrictedly} (many times).
%
As an example, consider from the following functions the first, which doesn't
typecheck because it is a linear function (annotated with the linearity
\texttt{1}) that consumes its argument more than once, and the second, which is
an unrestricted function (annoted with \texttt{Many}) that typechecks because
its type allows the argument to be consumed unrestrictedly.
%
\begin{code}
    f :: a %1 -> (a, a)
    f x = (x, x)
\end{code}
\begin{code}
    g :: a %Many -> (a, a)
    g x = (x, x)
\end{code}
%
The function annotated with the linearity annotation of \texttt{1} is equivalent
to the linear function type ($\lolli$) seen in the linear lambda
calculs~\ref{sec:linear-types}. 
%
Additionally, algebraic data type constructors can specify whether their
arguments are linear or unrestricted, requiring that, when pattern matched on,
linear arguments be consumed once while unrestricted arguments need not be
consumed exactly once.
%Similarly, when constructing such a constructor, using
% a linear value as a linear argument of the constructor consumes it once, while
% using it as an unrestricted argument consumes it unrestrictedly.
%
To encode the multiplicative linear pair ($\tensor$) we must create a pair data
type with two linear components. To consume an algebraic data type is to consume
all its linear components once, so to consume said pair data type we need to
consume both its linear components -- successfully encoding the
multiplicative pair elimination rule ($\tensor E$). To construct said pair data
type we must provide two linear elements, each consuming some required resources
to be constructed, thus encoding the multiplicative pair introduction rule
($\tensor I$). As such, we define \texttt{MultPair} as an algebraic data type
whose constructor uses a linear arrow for each of the arguments\footnote{By
default, constructors defined without GADT syntax have linear arguments. We
could have written \texttt{data MultPair a b = MkPair a b} to the same effect.}

@@

The linearity annotations are just a specialization of the more general so-called
\emph{multiplicity annotations}. A multiplicity of \texttt{1} entails that the
function argument must be consumed once, and a function annotated with it
($\to_1$) is called a linear function (also known as $\lolli$). A function with
multiplicity of \texttt{Many} ($\to_\omega$) allows the function argument to
be consumed unrestrictedly. Unrestricted functions are equivalent to the common
function type and, in fact, the usual function arrow ($\to$) implicitly has
multiplicity \texttt{Many}.  Multiplicities also smoothly allow for
\emph{multiplicity polymorphism}, which we describe below.

% TODO: Is this enough about multiplicities?

\begin{itemize}
    % \item Linear Haskell definition
    % \item Consuming values precisely
    % \item Multiplicities
    \item Multiplicity polymorphism
    \item Que relaciona o systemFC com o calculo lambda linear
\end{itemize}
